Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's Irises

### 2021-04-27 penguins.Rmd
Aph deux (#21)
* repurposed R/penguins_code.Rmd with penguins_plots.R to reduce duplication

* rm fig dir

* start keeping these files? not sure if we want to separate

* commit those files

* all png files now go here when you knit

* this is a purled script

* add tex file to git repo

* rework manuscript for better reproducibility

* don't show wizard behind curtain

* booktabs

* small edits

* updates

* no more gt?

* I don't even know

 In 1935, Edgar Anderson collected size measurements for 150 flowers from three species of *Iris* on the Gaspé Peninsula in Quebec, Canada. Since then, Anderson's *Iris* observations have become a classic dataset in statistics, machine learning, and data science teaching materials. It is included in the base R **datasets** package as **iris**, making it easy for users to access without knowing much about it. However, the lack of data documentation, presence of non-intuitive variables (e.g. "sepal width"), and perfectly balanced groups with zero missing values make **iris** an inadequate and stale dataset for teaching and learning modern data science skills. Users would benefit from working with a more representative, real-world environmental dataset with a clear link to current scientific research. Importantly, Anderson’s *Iris* dataset appeared in a 1936 publication by R. A. Fisher in the *Annals of Eugenics* (which is often the first-listed citation for the dataset), inextricably linking **iris** to racist research and theory. Thus, a modern alternative to the **iris** dataset is needed. 
 Here, we introduce the **palmerpenguins** R package, which includes morphological measurements collected from 2007 - 2009 for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica. The **penguins** data provides an approachable, charismatic, and near drop-in replacement for **iris** with topical relevance global climate change and environmental impacts on marine predators. Since being added to CRAN in July 2020, the **palmerpenguins** package has been downloaded over 102,000 times, highlighting the demand and widespread adoption of this viable **iris** alternative. 
 We directly compare the **iris** and **penguins** datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using **iris**, can switch to the Palmer Archipelago penguins for many use cases including data wrangling and visualization, correlation, regression, multivariate analysis (e.g., PCA), cluster analysis and classification (e.g., by k-means).

preamble: |
  % Any extra LaTeX you need in the preamble
output: rticles::rjournal_article
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(palmerpenguins)
library(kableExtra)
library(patchwork)
library(gt)
library(GGally)
library(recipes)
library(shadowtext)
library(broom)
knitr::opts_chunk$set(echo = TRUE)

theme_set(theme_minimal())
```

@@

In 1935, American botanist Edgar Anderson measured petal and sepal structural dimensions (length and width) for 50 flowers from three *Iris* species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson_irises_1935]. The manageable but non-trivial size (`r ncol(iris)` variables and `r nrow(iris)` total observations) and characteristics of the dataset, including linear relationships and multivariate normality, have made the dataset amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning. Anderson's *Iris* data is built into a number of software packages including the auto-installed datasets package in R [@r_core_team_r_2019], Python's scikit-learn machine learning library [@pedregosa_scikit-learn_2011], and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the *Iris* data. As a result, eighty-six years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics, computational methods, software documentation, and data science courses and materials.

There are a number of reasons that modern data science practitioners and educators may want to move on from Anderson's *Iris* dataset. First, Anderson's *Iris* dataset lacks metadata almost entirely [@anderson_irises_1935], which does not reinforce best practices and limits meaningful interpretation and discussion of research methods, analyses, and outcomes. Of the five variables in the dataset, two (`Sepal.Width` and `Sepal.Length`) are not intuitive for most non-botanists. Even with explanation, the difference between *Iris* *petal* and *sepal* dimensions is not obvious. Second, the *Iris* dataset contains no missing values and sample sizes for each of the three species (*n* = 50) are equal, which is cleaner than most real-world datasets that learners are likely to encounter "in the wild." Third, the single factor (`Species`) in the *Iris* data limits options for analyses. Finally, due to its publication in the *Annals of Eugenics* by eugenicist and statistician R.A. Fisher [@fisher_use_1936], the *Iris* dataset is burdened by a history in scientific racism, which we are committed to addressing through the development of new data science education products as described below.

@@

- Available by appropriate license (ideally, CC0 "no rights reserved")
- Feature intuitive subjects and variables that are interesting and understandable to learners across disciplines
- Complete metadata and documentation
- Manageable (but not trivial) in size
- Minimal data cleaning and pre-processing required for most analyses
- Real-world (not manufactured), modern data
- Provides similar opportunities for teaching and learning R, data science, and statistical skills
- Can easily replace *Iris* for most use cases

@@

Information, examples, and links to community-contributed materials are available on the \CRANpkg{palmerpenguins} package website: \url{https://allisonhorst.github.io/palmerpenguins/}.

@@

- *species:* a factor denoting the penguin species (Adélie, Chinstrap, or Gentoo)
- *island:* a factor denoting the island (Palmer Archipelago, Antarctica) where penguins were observed (Biscoe Point, Dream Island or Torgersen Island)
- *bill_length_mm:* a number denoting length of the dorsal ridge of a penguin bill (millimeters)
- *bill_depth_mm:* a number denoting the depth of a penguin bill (millimeters)
- *flipper_length_mm:* an integer denoting the length of a penguin flipper (millimeters)
- *body_mass_g:* an integer denoting the weight of a penguin’s body (grams)
- *sex:* a factor denoting the sex of a penguin sex (male, female) based on molecular data
- *year:* an integer denoting the year of study (2007, 2008 or 2009)

@@

```
    # install.packages("readr")
    library(palmerpenguins)
    df <- readr::read_csv(path_to_file("penguins.csv"))
```

@@

## Other data access options

## Software and code

- \CRANpkg{tidyverse} [@wickham_welcome_2019]: for data import and cleaning
- \CRANpkg{ggplot2} [@wickham_ggplot2_2016]: for data visualizations
- \CRANpkg{here} [@muller_here_2017]: for file path control
- \CRANpkg{kableExtra} [@zhu_kableextra_2019]: for finalized tables
- \CRANpkg{gt} [@iannone_gt_2020]: for finalized tables
- \CRANpkg{GGally} [@schloerke_ggally_2020]: for pairs plots
- \CRANpkg{patchwork} [@pedersen_patchwork_2019]: for compound figures
- \CRANpkg{shadowtext} [@yu_shadowtext_2019]: to add a background color to text labels
- \CRANpkg{recipes} [@kuhn_recipes_2020]: for data pre-processing
- \CRANpkg{pkgdown} [@wickham_pkgdown_2020]: to build the package website
- \pkg{base} and \pkg{stats} [@r_core_team_r_2019]: for various analyses throughout

## Selected comparisons between **iris** and **penguins**

@@

### Overview

```{r, echo = FALSE}
overview_compare <- tribble(
  ~Feature, ~iris, ~penguins,
  "Year(s) collected", "1935", "2007 - 2009",
  "Dimensions (col x row)", "5 x 150", "8 x 344",
  "Documentation", "minimal", "complete metadata",
  "Variable classes", "double (4), factor (1)", "double (2), int (3), factor (3)",
  "Missing values?", "no (n = 0; 0.0%)", "yes (n = 19; 0.7%)"
)

overview_compare %>% 
  kable(caption = "Overview comparison of penguins and iris dataset features and characteristics.")
```

@@

```{r, echo = FALSE}
iris_counts <- iris %>% 
  count(Species) %>% 
  rename(`Iris species` = Species)

@@

iris_penguin_n %>% 
   kable(caption = "Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total). Penguins can be further grouped by variables for island and study year.", align = "lclccc") %>% 
  add_header_above(c("Iris sample size (by species)" = 2, "Penguin sample size (by species and sex)" = 4 ))

@@

Distributions, relationships between variables, and clustering can be visually explored between species for the four structural size measurements in **penguins** (flipper length, body mass, bill length and depth; Figure \ref{fig:penguins-pairs}) and **iris** (sepal width and length, petal width and length; Figure \ref{fig:iris-pairs}).

```{r, penguins-pairs, out.width = "100%", fig.cap = "Distributions and correlations for numeric variables in the penguins data (flipper length (mm), body mass (g), bill length (mm) and bill depth (mm)) for the three observed species: Gentoo (green, triangles); Chinstrap (blue, circles); and Adélie (orange, squares). Significance levels indicated for bivariate correlations are: *p < 0.05; **p < 0.01; ***p < 0.001.", warning = FALSE, message = FALSE, echo = FALSE}

knitr::include_graphics(here("fig", "penguin_pairs.png"))

```{r, iris-pairs, out.width = "100%", fig.cap = "Distributions and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares). Significance levels indicated for bivariate correlations are: *p < 0.05; **p < 0.01; ***p < 0.001.", warning = FALSE, message = FALSE, echo = FALSE}

knitr::include_graphics(here("fig", "iris_pairs.png"))

Both **penguins** and **iris** offer numerous opportunities to explore linear relationships and correlations, within and across species (Figures \ref{fig:penguins-pairs} & \ref{fig:iris-pairs}). A bivariate scatterplot made with the **iris** dataset reveals a clear linear relationship between petal length and petal width. Using **penguins** (\ref{fig:linear-example}), we can create a uniquely similar scatterplot with flipper length and body mass. The overall trend across all three species is approximately linear for both **iris** and **penguins**. Teachers may encourage students to explore how simple linear regression results and predictions differ when the species variable is omitted, compared to multiple linear regression with species included (\ref{fig:linear-example}).

```{r, linear-example, out.width = "100%", fig.cap = "Representative linear relationships for (A): penguin flipper length (mm) and body mass (g) for Adélie (orange circles), Chinstrap (blue triangles), and Gentoo (green squares) penguins; (B): iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises. Within-species linear model is visualized for each penguin or iris species.", echo = FALSE, fig.align='center', fig.pos='htbp'}

knitr::include_graphics('fig/linear_example.png')

@@

Simpson’s Paradox is a data phenomenon in which a trend observed between variables is reversed when data are pooled, omitting a meaningful variable. While often taught and discussed in statistics courses, finding a real-world and approachable example of Simpson’s Paradox can be a challenge. Here, we show one (of several possible - see Figure \ref{fig:penguins-pairs}) Simpson’s Paradox examples in **penguins**: exploring bill dimensions with and without species included (Figure \ref{fig:simpsons}). When penguin species is omitted (Figure \ref{fig:simpsons} A), bill length and depth appear negatively correlated overall. The trend is reversed when species is included, revealing an obviously positive correlation between bill length and bill depth within species (Figure \ref{fig:simpsons} B).

```{r, simpsons, out.width = "100%", fig.cap = "Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the species variable is excluded (A) or included (B), illustrating Simpson’s Paradox. Note: linear regression for bill dimensions without including species in (A) is ill-advised; the linear trendline is only included to visualize trend reversal for Simpson’s Paradox when compared to (B).", echo = FALSE, fig.align='center', fig.pos='htbp'}

knitr::include_graphics('fig/simpson_gg.png')

@@

```{r, pca, out.width = "100%", fig.cap = "Principal component analysis biplots and screeplots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component. For penguins, variables are flipper length (mm), body mass (g), bill length (mm) and bill depth (mm); groups are visualized by species (Adélie = orange circles, Chinstrap = blue triangles, Gentoo = green squares). For iris, variables are petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm); groups are visualized by species (setosa = light gray circles, versicolor = dark gray triangles, virginica = black squares). Values above screeplot columns (C,D) indicate percent of total variance explained by each of the four principal components.", echo = FALSE, fig.align='center', fig.pos='htbp'}

knitr::include_graphics('fig/pca_plots.png')

@@

```{r, include = FALSE, echo = FALSE}
# Included to create kable table of cluster assignment
# k-means comparison
# Penguins: Bill length vs. bill depth
# Iris: petal length vs. petal width

@@

# Making cluster assignments table
kmeans_2var_table <- cbind(pb_clust_n, ip_clust_n) %>% 
  kable(col.names = c("Cluster", "Adélie", "Chinstrap", "Gentoo", "Cluster", "setosa", "versicolor", "virginica"),
        caption = "K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).", align = "cccccc") %>% 
  kable_styling(full_width = FALSE) %>% 
  add_header_above(c("Penguins cluster assignments" = 4, "Iris cluster assignments" = 4))
```

```{r, kmeans, out.width = "100%", fig.cap = "K-means clustering outcomes for penguin bill dimensions (A) and iris petal dimensions (B). Numbers indicate the cluster to which an observation was assigned, revealing a high degree of separation between species for both penguins and iris.", echo = FALSE, fig.align='center', fig.pos='htbp'}

knitr::include_graphics('fig/kmeans.png')
```

```{r, echo = FALSE}
kmeans_2var_table
```

@@

## Appendix

### Appendix A

@@

### Appendix B

Summary table, **penguins_raw** dataset in the **palmerpenguins** package.

@@

overview_penguins_raw %>% 
  kable()
```

<!---This section may contain a figure such as Figure \ref{fig:Rlogo}.

```{r, Rlogo, echo=FALSE, fig.cap='The logo of R.', out.width='2in', fig.align='center', fig.pos='htbp'}
knitr::include_graphics('palmerpenguins-logo.png')
``` 
--->

<!---
## Another section

There will likely be several sections, perhaps including code snippets, such as:

```{r}
x <- 1:10
x
```
--->

<!----
## Summary

This file is only a basic article template. For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf).

--->

>>>>

 In 1935, Edgar Anderson collected size measurements for 150 flowers from three species of *Iris* on the Gaspé Peninsula in Quebec, Canada. Since then, Anderson's *Iris* observations have become a classic dataset in statistics, machine learning, and data science teaching materials. It is included in the base R **datasets** package as **iris**, making it easy for users to access without knowing much about it. However, the lack of data documentation, presence of non-intuitive variables (e.g. "sepal width"), and perfectly balanced groups with zero missing values make **iris** an inadequate and stale dataset for teaching and learning modern data science skills. Users would benefit from working with a more representative, real-world environmental dataset with a clear link to current scientific research. Importantly, Anderson’s *Iris* dataset appeared in a 1936 publication by R. A. Fisher in the *Annals of Eugenics* (which is often the first-listed citation for the dataset), inextricably linking **iris** to racist research and theory. Thus, a modern alternative to the **iris** dataset is needed. In this paper, we introduce the **palmerpenguins** R package, which includes morphological measurements collected from 2007 - 2009 for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica. The **penguins** data provides an approachable, charismatic, and near drop-in replacement for **iris** with topical relevance global climate change and environmental impacts on marine predators. Since the release on CRAN in July 2020, the **palmerpenguins** package has been downloaded over 102,000 times, highlighting the demand and widespread adoption of this viable **iris** alternative. 
 We directly compare the **iris** and **penguins** datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using **iris**, can switch to the Palmer Archipelago penguins for many use cases including data wrangling, visualization, linear modeling, multivariate analysis (e.g., PCA), cluster analysis and classification (e.g., by k-means).

preamble: |
  % Any extra LaTeX you need in the preamble
output: rticles::rjournal_article
bibiography: packages.bib
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      purl = TRUE,
                      out.width = "100%",
                      dpi = 300, 
                      fig.align = "center",
                      fig.path = "figs/",
                      dev = "png")
```

```{r pkgs, include=FALSE, purl=TRUE}
library(tidyverse)
library(palmerpenguins)

# for tables
library(kableExtra)

# for plots
library(paletteer)
library(patchwork)
library(GGally)
library(shadowtext)
ggplot2::theme_set(ggplot2::theme_minimal())

# for modeling
library(recipes)
library(broom)

knitr::read_chunk("penguins_plots.R")
```

```{r write-bib, purl=FALSE, include=FALSE}
knitr::write_bib(.packages(), "packages.bib")
```

@@

In 1935, American botanist Edgar Anderson measured petal and sepal structural dimensions (length and width) for 50 flowers from three *Iris* species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson_irises_1935]. The manageable but non-trivial size (`r ncol(iris)` variables and `r nrow(iris)` total observations) and characteristics of the dataset, including linear relationships and multivariate normality, have made the dataset amenable for introducing a wide range of statistical methods including data wrangling, visualization, linear modeling, multivariate analyses, and machine learning. Anderson's *Iris* data is built into a number of software packages including the auto-installed datasets package in R [@r_core_team_r_2019], Python's scikit-learn machine learning library [@pedregosa_scikit-learn_2011], and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the *Iris* data. As a result, eighty-six years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics, computational methods, software documentation, and data science courses and materials.

There are a number of reasons that modern data science practitioners and educators may want to move on from Anderson's *Iris* dataset. First, Anderson's *Iris* dataset lacks metadata almost entirely [@anderson_irises_1935], which does not reinforce best practices and limits meaningful interpretation and discussion of research methods, analyses, and outcomes. Of the five variables in the dataset, two (`Sepal.Width` and `Sepal.Length`) are not intuitive for most non-botanists. Even with explanation, the difference between *Iris* *petal* and *sepal* dimensions is not obvious. Second, the *Iris* dataset contains equal sample sizes for each of the three species (*n* = 50) with no missing values, which is cleaner than most real-world datasets that learners are likely to encounter "in the wild." Third, the single factor (`Species`) in the *Iris* data limits options for analyses. Finally, due to its publication in the *Annals of Eugenics* by eugenicist and statistician R.A. Fisher [@fisher_use_1936], the *Iris* dataset is burdened by a history in scientific racism, which we are committed to addressing through the development of new data science education products as described below.

@@

- Available by appropriate license (ideally, CC0 "no rights reserved")
- Feature intuitive subjects and variables that are interesting and understandable to learners across disciplines
- Complete metadata and documentation
- Manageable (but not trivial) in size
- Minimal data cleaning and pre-processing required for most analyses
- Real-world (not manufactured) modern data
- Provides similar opportunities for teaching and learning R, data science, and statistical skills
- Can easily replace *Iris* for most use cases

@@

Information, examples, and links to community-contributed materials are available on the \CRANpkg{palmerpenguins} package website: \url{https://allisonhorst.github.io/palmerpenguins/}. See the Appendix for how Python and Julia users can access the same data.

@@

- *species:* a factor denoting the penguin species (Adélie, Chinstrap, or Gentoo)
- *island:* a factor denoting the Palmer Archipelago island in Antarctica where each penguin was observed (Biscoe Point, Dream Island, or Torgersen Island)
- *bill_length_mm:* a number denoting length of the dorsal ridge of a penguin bill (millimeters)
- *bill_depth_mm:* a number denoting the depth of a penguin bill (millimeters)
- *flipper_length_mm:* an integer denoting the length of a penguin flipper (millimeters)
- *body_mass_g:* an integer denoting the weight of a penguin’s body (grams)
- *sex:* a factor denoting the sex of a penguin sex (male, female) based on molecular data
- *year:* an integer denoting the year of study (2007, 2008, or 2009)

\newpage

@@

```
    library(palmerpenguins)
    df <- read.csv(path_to_file("penguins.csv"))
```

@@

## Comparing **iris** and **penguins**

@@

```{r overview-tbl}
overview_compare <- tribble(
  ~Feature, ~iris, ~penguins,
  "Year(s) collected", "1935", "2007 - 2009",
  "Dimensions (col x row)", "5 x 150", "8 x 344",
  "Documentation", "minimal", "complete metadata",
  "Variable classes", "double (4), factor (1)", "double (2), int (3), factor (3)",
  "Missing values?", "no (n = 0; 0.0%)", "yes (n = 19; 0.7%)"
)

overview_compare %>% 
  kable(caption = "Overview comparison of penguins and iris dataset features and characteristics.",
        align = "lclccc",
        booktabs = TRUE)
```

@@

```{r counts-tbl}
iris_counts <- iris %>% 
  count(Species) %>% 
  rename(`Iris species` = Species)

@@

iris_penguin_n %>% 
   kable(caption = "Grouped sample size for iris (by species; n = 150 total) and penguins (by species and sex; n = 344 total). Penguins can be further grouped by variables for island and study year.", 
         align = "lclccc",
         booktabs = TRUE) %>% 
  add_header_above(c("Iris sample size (by species)" = 2, "Penguin sample size (by species and sex)" = 4 ))

@@

Distributions, relationships between variables, and clustering can be visually explored between species for the four structural size measurements in **penguins** (flipper length, body mass, bill length and depth; Figure \ref{fig:penguin-pairs}) and **iris** (sepal width and length, petal width and length; Figure \ref{fig:iris-pairs}).

```{r, penguin-pairs, width = 6, height = 4.5, fig.cap = "Distributions and correlations for numeric variables in the penguins data (flipper length (mm), body mass (g), bill length (mm) and bill depth (mm)) for the three observed species: Gentoo (green, triangles); Chinstrap (blue, circles); and Adélie (orange, squares). Significance levels indicated for bivariate correlations are: *p < 0.05; **p < 0.01; ***p < 0.001."}
```

```{r, iris-pairs, width = 6, height = 4.5, fig.cap = "Distributions and correlations for numeric variables in the iris data (petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm)) for the three included Iris species: setosa (light gray, circles); versicolor (dark gray, triangles); and virginica (black, squares). Significance levels indicated for bivariate correlations are: *p < 0.05; **p < 0.01; ***p < 0.001."}
```

Both **penguins** and **iris** offer numerous opportunities to explore linear relationships and correlations, within and across species (Figures \ref{fig:penguin-pairs} & \ref{fig:iris-pairs}). A bivariate scatterplot made with the **iris** dataset reveals a clear linear relationship between petal length and petal width. Using **penguins** (\ref{fig:linear-example}), we can create a uniquely similar scatterplot with flipper length and body mass. The overall trend across all three species is approximately linear for both **iris** and **penguins**. Teachers may encourage students to explore how simple linear regression results and predictions differ when the species variable is omitted, compared to multiple linear regression with species included (\ref{fig:linear-example}).

```{r, linear-example, width = 6, height = 3, fig.cap = "Representative linear relationships for (A): penguin flipper length (mm) and body mass (g) for Adélie (orange circles), Chinstrap (blue triangles), and Gentoo (green squares) penguins; (B): iris petal length (cm) and width (cm) for setosa (light gray circles), versicolor (dark gray triangles) and virginica (black squares) irises. Within-species linear model is visualized for each penguin or iris species.", fig.pos='htbp'}
```

@@

Simpson’s Paradox is a data phenomenon in which a trend observed between variables is reversed when data are pooled, omitting a meaningful variable. While often taught and discussed in statistics courses, finding a real-world and approachable example of Simpson’s Paradox can be a challenge. Here, we show one (of several possible - see Figure \ref{fig:penguin-pairs}) Simpson’s Paradox examples in **penguins**: exploring bill dimensions with and without species included (Figure \ref{fig:simpsons}). When penguin species is omitted (Figure \ref{fig:simpsons} A), bill length and depth appear negatively correlated overall. The trend is reversed when species is included, revealing an obviously positive correlation between bill length and bill depth within species (Figure \ref{fig:simpsons} B).

```{r, simpsons, width = 6, height = 2.5, fig.cap = "Trends for penguin bill dimensions (bill length and bill depth, millimeters) if the species variable is excluded (A) or included (B), illustrating Simpson’s Paradox. Note: linear regression for bill dimensions without including species in (A) is ill-advised; the linear trendline is only included to visualize trend reversal for Simpson’s Paradox when compared to (B).", fig.pos='htbp'}
```

@@

```{r, pca, width = 8, height = 8, fig.cap = "Principal component analysis biplots and screeplots for structural size measurements in penguins (A,C) and iris (B,D), revealing similarities in multivariate patterns, variable loadings, and variance explained by each component. For penguins, variables are flipper length (mm), body mass (g), bill length (mm) and bill depth (mm); groups are visualized by species (Adélie = orange circles, Chinstrap = blue triangles, Gentoo = green squares). For iris, variables are petal length (cm), petal width (cm), sepal length (cm) and sepal width (cm); groups are visualized by species (setosa = light gray circles, versicolor = dark gray triangles, virginica = black squares). Values above screeplot columns (C,D) indicate percent of total variance explained by each of the four principal components.", fig.pos='htbp'}
```

@@

```{r}
# Included to create kable table of cluster assignment
# k-means comparison
# Penguins: Bill length vs. bill depth
# Iris: petal length vs. petal width

@@

# Making cluster assignments table
kmeans_2var_table <- cbind(pb_clust_n, ip_clust_n) %>% 
  kable(col.names = c("Cluster", "Adélie", "Chinstrap", "Gentoo", "Cluster", "setosa", "versicolor", "virginica"),
        caption = "K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).", 
        align = "cccccc",
        booktabs = TRUE) %>% 
  kable_styling(full_width = FALSE) %>% 
  add_header_above(c("Penguins cluster assignments" = 4, "Iris cluster assignments" = 4))
```

```{r, kmeans, width = 8, height = 4.5, fig.cap = "K-means clustering outcomes for penguin bill dimensions (A) and iris petal dimensions (B). Numbers indicate the cluster to which an observation was assigned, revealing a high degree of separation between species for both penguins and iris.", fig.pos='htbp'}
```

```{r}
kmeans_2var_table
```

@@

# Appendix {-}

## Penguins data processing

@@

## Summary of the **penguins_raw** dataset

@@

overview_penguins_raw %>% 
  kable(booktabs = TRUE)
```

## Other data sources

## Acknowledgements

- \CRANpkg{tidyverse} [@wickham_welcome_2019]: for data import and cleaning
- \CRANpkg{ggplot2} [@wickham_ggplot2_2016]: for data visualizations
- \CRANpkg{kableExtra} [@zhu_kableextra_2019]: for finalized tables
- \CRANpkg{GGally} [@schloerke_ggally_2020]: for pairs plots
- \CRANpkg{patchwork} [@pedersen_patchwork_2019]: for compound figures
- \CRANpkg{shadowtext} [@yu_shadowtext_2019]: to add a background color to text labels
- \CRANpkg{recipes} [@kuhn_recipes_2020]: for data pre-processing
- \CRANpkg{pkgdown} [@wickham_pkgdown_2020]: to build the package website
- \pkg{base} and \pkg{stats} [@r_core_team_r_2019]: for various analyses throughout
